Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude

### 2018-06-26 paper/paper_after_guide.tex
added appendix and one way transformation to the paper

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{array}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{ebproof}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\author{ANDREI-ALIN CORODESCU}
\title{Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude}
\newcommand{\hq}[4]{
	\left\{{#1}\right\}
	\let\scriptstyle\textstyle 
	\substack{{#2} \\ 	\let\scriptstyle\textstyle {#3}} 
	\left\{{#4}\right\}
}

@@

\lstset{style=mystyle}
\begin{document}
\begin{titlepage}
\begin{center}
\textbf{
ALEXANDRU IOAN CUZA UNIVERSITY OF IAȘI
}
\\
\textbf{FACULTY OF COMPUTER SCIENCE}
\end{center}
   \vspace{40mm}
\begin{center}
	\Large\textbf {Program Equivalence: An Interactive Relational Separation Logic Prover Implemented in Maude}\\
	\vspace{40mm}
	\large\textit {ANDREI-ALIN CORODESCU}
	\\
	\vspace{20mm}
	\textbf{Session: }\textit{July, 2018}\\
	\vspace{30mm}
	\textbf{Scientific Coordinator}\\
	\textbf{\textit{Conf. Dr. Ciobâcă Ștefan}}
	\vspace{30mm}
\end{center}
\end{titlepage}
\begin{abstract}
	Lucrarea de fata descrie procesul de dezvoltare a unui utilitar interactiv utilizat in analiza formala a programelor, folosindu-se de Logica Separatoare Relationala \cite{relational} si Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizand Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactiva se imbina cu o componenta automatizata, prin care parti din demonstratie vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul functionalitatilor oferite de Maude si va include secvente de cod ce vor ilustra conceptele cheie in modelarea si aplicarea fundamentelor teoretice.
\end{abstract}
\pagebreak
\tableofcontents
\pagebreak
\section*{Introduction}
The present paper describes the development of an interactive tool for reasoning how two programs are related, based theoretical concepts from the field of formal program analysis and technologies which facilitate the implementation of those concepts. \\

@@

The rest of the paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} will describe the main personal contributions to the realization of this project}
	\item {\textbf{Section 2} will present the problem this project aims to solve}
	\item {\textbf{Section 3} will shortly present some other projects related to ours}
	\item {\textbf{Section 4} will briefly touch upon the theoretical foundations of this project, together with the technologies used to implement them}
	\item {\textbf{Section 5}, the main section of this paper, will present our approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\pagebreak
\section{Contributions}
Personal contributions to the realization of the project : 
\begin{itemize}
	\item Modelled the Relational Logic and Separation Logic using Maude equational and rewriting logic specifications . 
	\item Developed an interactive tool for reasoning about program behaviour using the aforementioned logics.
	\item Automation of tasks, which makes the tool more convenient to use .
	\item Example of formal proof done using the tool \textbf{need annex}
\end{itemize}
\section{Description of the problem}
The problem this project is aiming to solve is related to formal reasoning about the execution of code, mainly focusing on how two programs are related to each other (most often the relation to be proven is equivalence)\\

@@

\section{Theoretical Foundations and technologies}
This section will briefly introduce the concepts on which the project is based, together with the technologies used to implement it, along with a few references to resources which explain them in depth.
\subsection{Separation Logic}
Separation Logic allows for reasoning about the effects of code on the program state in a formal manner. The main abstraction used in separation logic is the \textbf{Hoare Triple}
\[ \{P\}\  C\  \{Q\} \]
where \(P\) denotes the a condition satisfied by state of the program before the execution of the command \(C\) and \(Q\) denotes another condition satisfied by state of the program after the execution of \(C\).
\\

@@

	\paragraph{Executable semantics of the language, in Maude}
	Maude also supports specifying executable semantics for the language using rewriting logic \cite{rewrConcurrency}. A proof of concept for a language similar to the one used by the prover is present in the code of the project, but it is out of scope for the presentation of the prover \textbf{Need reference to the final file name}. An executable environment for the programming language used in the prover is a valid candidate for future improvements.

@@

	\[
		[\mathtt{INPUT},\ \mathtt{PROVER-STATE},\ \mathtt{OUPTUT}]
	\]
	\label{fig:loopState}
	\caption{Loop state}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}
\subsection{State of the program}
Following the examples provided together with the Maude Manual \cite{manual}, our state (Figure \ref{fig:ProverState}) is composed multiple sections modified accordingly during program execution:
\begin{itemize}
	\item{\textbf{Action} - represents the current action to be executed by the prover, more details about supported actions are included in the following sections.}
	\item{\textbf{Root Goal} - represents the end-goal of this demonstration}
	\item{\textbf{Current Goal} - the goal being treated currently - all actions will be applied to this goal}
	\item{\textbf{Goal Stack} - Stack of goals necessary to be proven in order to prove the root goal. The current goal is the last element popped from the stack and new elements are generated and pushed by the application of proof rules to the current goal.}
	\item{\textbf{Proven Goals} - List of goals which have been already been proven or axioms. Each goal will be first checked to see if it matches any goal in this list to prevent repeating the same proofs.}
	\item{\textbf{Staging output} - contains the output generated by the current action, before being passed to the Loop state. This staging mechanism enables each rewriting rule to generate output independent of the others.}
\end{itemize}
\begin{figure}[h]
	\noindent\rule{\linewidth}{0.4pt}
	\(
	< \mathtt{Action}, \mathtt{Root Goal}, \mathtt{Current Goal}, \mathtt{Goal Stack}, \mathtt{Proven Goals}, \mathtt{Staging Output} >
	\)
	\caption{Prover State}
	\label{fig:ProverState}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}
\subsection{Input and output handling}
To understand how our prover handles input and output, it is important to note that the flow of the LOOP-MODE module is as follows: 
\begin{enumerate}
	\item User input is placed in the \(\mathtt{INPUT}\) member of the loop state
	\item Maude engine applies all the rewritings possible, resulting in a new loop state.

@@

\subsection{Prover execution flow}
In this section we will describe the flow of a demonstration made using our prover. The most important parts of the prover's execution are controlled through the first member of a prover state, \textbf{Action}, which specifies the next step in the processing of the current goal. There are two types of actions, \textbf{System Actions}, colored in the figure in purple and \textbf{User Actions}, colored in light-yellow. The diagram presenting how the prover works is presented in Figure \ref{fig:ProverDiagram}
\\
\begin{enumerate}
	\item The user issues the command \texttt{loop init .} in the Maude console. This will initialize the loop state.
	\item Now the user will have to specify the goal it wants to prove, by issuing the command \begin{center}(\texttt{prove(G)})\end{center}, where G is the representation of the goal using the syntax defined for our logics. The command will populate the required fields of the prover state (\textbf{RootGoal}, \textbf{GoalStack}) with G.
	\\

	This command will also populate the \textbf{ProvenGoals} list with axioms.
	\\

@@

	\item The next phase involves applying the logic proof rules to the goals in the \textbf{GoalStack} to generate sub-goals and prove them. Once a goal is popped from the \textbf{GoalStack}, it can be replaced by sub-goals generated by proof rules, matched with an proven goal / axiom, be proven automatically or be admitted manually. This phase lasts until there are no more goals left in the \textbf{GoalStack}, meaning that the \textbf{RootGoal} has been proven.

@@

\paragraph{Goal Graph} One of the improvements which could make the user experience much better for our prover would be the capability to backtrack in the proof and try another approach for the same goal. The initial idea for this prover was to have a \textbf{GoalGraph} instead of \textbf{GoalStack} such that when the user applies a rule to a goal, the sub goals are represented as nodes in the same graph, making it possible to trace back the progress to intermediary states. This approach would also make the \textbf{ProvenGoals} a lot more useful by increasing the chance of running into an already proven goal and the ability to mark complex goals as proven. At the moment, the only goals which can be used from the \textbf{ProvenGoals} are those admitted manually by the user, since the other proven goals already matched axioms.

@@

		[\textbf{SL}] + \textbf{apply} + \textit{RuleName}

@@

		With \textbf{SL} being added only when applying Separation Logic rules, to distinguish between the Relational Separation Logic rules with the same name.

@@

			\item If the result corresponds to what the prover was expecting, we modify the state by adding the newly generated goals to the \textbf{GoalStack}, by moving down the "meta"-hierarchy - parsing meta-representation of terms.
			\item The last thing to note is an example of how the rules generate output independent of other applications. In the last field of the \textbf{Prover State}, as \textbf{StagingOutput}, there will be a list of quoted identifiers which will eventually get moved to the loop state and then outputted to the user.

@@

	\begin{lstlisting}[label=axiom,caption=Initialization of Prover state]
rl [proveGoal] : < prove(HQ) ; RG ; G ; GS ; GL ; nil > =>
< noAction ; HQ ; noGoal ; HQ <> emptyGL ; 
downTerm('`{_`}_`{_`}['_|->_['E:AExp,'_:AExp],'`[_`]:=_['E:AExp,'F:AExp],'_|->_['E:AExp,'F:AExp]] , emptyGL)
--- ... [other axioms omitted for clarity] 
; 'begin '\n > .\end{lstlisting}
	\item Before attempting to match terms we also interpret the equality between variables using the \texttt{replace} operator introduced previously to reduce the \texttt{Goal} to a simpler form, which might match to an axiom.
	\begin{lstlisting}[caption=Rule intepreting variable equality]
rl {P * (emp /\ X eqs Y)} X1 := E {Q} =>  
if X1 =/= X then 
{replace(P, X, Y)} replace(X1 := E, X, Y) {replace(Q, X, Y)}
else 
{replace(P, X, Y)} X1 := replace(E, X, Y) {Q}
fi .\end{lstlisting}
\end{itemize}
\subsubsection{Automatic demonstration of implications}
Automatic demonstration of implications is one of the most challenging aspects of this paper and this section will present our incipient approach to this problem.

We have modeled implication proof rules as rewrite rules in Maude (example in Listing \ref{implicationProofRule}), in order to make use of its \texttt{metaSearchPath} operator, which will search for a series of rewrites which transform a term into another. We used this functionality to try and solve implications 
\[
	R \Rightarrow S
\]
by searching for a series of rewrites from \texttt{R => S} to \texttt{true}. 
\begin{lstlisting}[label=implicationProofRule,caption=Example of implication rewrite rules]
rl R:Relation * ( S:Relation \/ T:Relation ) => (R:Relation * S:Relation) \/ (R:Relation * T:Relation) .
--- Example for the usage of the replace operator
rl R * (Emp /\ X eqs Y) => replace(R, X , Y) .
rl (R:Relation => R:Relation) => true .
\end{lstlisting}

@@

The second parameter controls the depth up to which the search goes. The prover will try automatically to search a solution within a depth of 3 rewrites, but the user can manually specify searching for a solution using the \texttt{auto x} command, where \texttt{x} is the depth up to which Maude should try to search the solution (our experiments show that searches up to a depth around 12 should finish use a reasonable amount of resources and finish in under 10 minutes - this observation of course depends on the implication and the number of possible rewrites starting from it). The found path is presented to the user if the implication has been proven. 
\paragraph{Inductive predicates} - the prover currently supports defining inductive predicates only by modifying the underlying Maude code. We have defined a single predicate, used in the Relational Separation Logic paper \cite{relational} :
\begin{lstlisting}[caption=Predicate syntax definition]
op List : Id -> Relation . 
\end{lstlisting}
In terms of the semantics of the \texttt{List} predicate, we have added rules to the \textbf{implication-prover.maude} file, to be considered when rewriting implications.
\begin{lstlisting}[caption=Example of rule interpreting the List predicate during implication proving]
rl [ (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ; (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ] * List(Y:Id) => List(X) .
\end{lstlisting}
\paragraph{Cyclist}\cite{cyclist} \cite{cyclistSite} is an automatic entailment prover for Separation Logic which we planned to integrate with our prover for Separation Logic implications and eventually extend to Relational Separation Logic, since it provides a generic framework. Since Maude does not easily support external commands, the idea would be to modify Maude's C++ code and add a hook for the \texttt{[special]} operator attribute which would invoke that hook, which will call the external command for Cyclist. We have however focused on other aspects of our prover, and left this feature for the future.
\subsection{User Interface}
Since our project aims to simplify the demonstrations of Separation Logic / Relational Separation Logic formulae, using an interactive prover, the user interface plays a major role in the success of the project - the easier it is to use and understand, the more chances of success it has. This idea has been presented in \cite{primer}, when discussing key elements of tools related to program verification. \\

@@

Maude has been interesting to work because it required an entirely different mindset than the mainstream programming languages. Once grasping the concepts around Maude which was built, it was easy to come up with simple, clever solutions to the problems I was facing. \\
As it is described on it's website \cite{maudesite}, Maude is a powerful logical framework which allows modeling of other logics in it; it proved to be exactly that, enabling us to represent the logics in a natural manner, which made the implementation much less error-prone and allowed us to focus on the relevant part of this project, namely how to apply the logics instead of focusing on how to properly implement them.
\\

@@

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{array}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{ebproof}
\usepackage[toc,page]{appendix}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{afterpage}

\newcommand\blankpage{%
	\null
	\thispagestyle{empty}%
	\addtocounter{page}{-1}%
	\newpage}
\author{ANDREI-ALIN CORODESCU}
\title{Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude}
\newcommand{\hq}[4]{
	\left\{{#1}\right\}
	\let\scriptstyle\textstyle 
	\substack{{#2} \\ 	\let\scriptstyle\textstyle {#3}} 
	\left\{{#4}\right\}
}

@@

\lstset{style=mystyle}
\begin{document}
\begin{titlepage}
\begin{center}
\afterpage{\blankpage}
\textbf{
ALEXANDRU IOAN CUZA UNIVERSITY OF IAȘI
}
\\
\textbf{FACULTY OF COMPUTER SCIENCE}
\end{center}
   \vspace{40mm}
\begin{center}
	\Large\textbf {Program Equivalence: An Interactive Relational Separation Logic Prover Implemented in Maude}\\
	\vspace{40mm}
	\large\textit {ANDREI-ALIN CORODESCU}
	\\
	\vspace{20mm}
	\textbf{Session: }\textit{July, 2018}\\
	\vspace{30mm}
	\textbf{Scientific Coordinator}\\
	\textbf{\textit{Conf. Dr. Ciobâcă Ștefan}}
	\vspace{30mm}
\end{center}
\end{titlepage}

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare a unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactiva se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
The present paper describes the development of an interactive tool for reasoning how two programs are related, based theoretical concepts from the field of formal program analysis and technologies which facilitate the implementation of those concepts. \\

@@

The rest of the paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} will describe the main personal contributions to the realization of this project}
	\item {\textbf{Section 2} will present the problem this project aims to solve}
	\item {\textbf{Section 3} will shortly present some other projects related to ours}
	\item {\textbf{Section 4} will briefly touch upon the theoretical foundations of this project, together with the technologies used to implement them}
	\item {\textbf{Section 5}, the main section of this paper, will present our approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\section{Contributions}
Personal contributions to the realization of the project : 
\begin{itemize}
	\item Modelled the Relational Logic and Separation Logic using Maude equational and rewriting logic specifications . 
	\item Developed an interactive tool for reasoning about program behaviour using the aforementioned logics.
	\item Automation of tasks, which makes the tool more convenient to use .
	\item Example of formal proof done using the tool - Appendix \ref{proofExample}
\end{itemize}
\section{Description of the problem}
The problem this project is aiming to solve is related to formal reasoning about the execution of code, mainly focusing on how two programs are related to each other (most often the relation to be proven is equivalence)\\

@@

\section{Theoretical Foundations and technologies}
This section will briefly introduce the concepts on which the project is based, together with the technologies used to implement it, along with a few references to resources which explain them in depth.
\subsection{Separation Logic}
Separation Logic allows for reasoning about the effects of code on the program state in a formal manner. The main abstraction used in separation logic is the \textbf{Hoare Triple}
\[ \{P\}\  C\  \{Q\} \]
where \(P\) denotes the a condition satisfied by state of the prover before the execution of the command \(C\) and \(Q\) denotes another condition satisfied by state of the prover after the execution of \(C\).
\\

@@

	\paragraph{Executable semantics of the language, in Maude}
	Maude also supports specifying executable semantics for the language using rewriting logic \cite{rewrConcurrency}. A proof of concept for a language similar to the one used by the prover is present in the code of the project (\textbf{simpleLanguage.maude}), but it is out of scope for the presentation of the prover . An executable environment for the programming language used in the prover is a valid candidate for future improvements.

@@

	\[
		[\mathtt{INPUT},\ \mathtt{PROVER-STATE},\ \mathtt{OUPTUT}]
	\]
	\label{fig:loopState}
	\caption{Loop state}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}
\subsection{State of the Prover}
Following the examples provided together with the Maude Manual \cite{manual}, our state (Figure \ref{fig:ProverState}) is composed multiple sections modified accordingly during program execution:
\begin{itemize}
	\item{\texttt{Action} - represents the current action to be executed by the prover, more details about supported actions are included in the following sections.}
	\item{\texttt{Root Goal} - represents the end-goal of this demonstration}
	\item{\texttt{Current Goal} - the goal being treated currently - all actions will be applied to this goal}
	\item{\texttt{Goal Stack} - Stack of goals necessary to be proven in order to prove the root goal. The current goal is the last element popped from the stack and new elements are generated and pushed by the application of proof rules to the current goal.}
	\item{\texttt{Proven Goals} - List of goals which have been already been proven or axioms. Each goal will be first checked to see if it matches any goal in this list to prevent repeating the same proofs.}
	\item{\texttt{Staging Output} - contains the output generated by the current action, before being passed to the Loop state. This staging mechanism enables each rewriting rule to generate output independent of the others.}
\end{itemize}
\begin{figure}[h]
	\noindent\rule{\linewidth}{0.4pt}
	\(
	< \mathtt{Action}, \mathtt{Root Goal}, \mathtt{Current Goal}, \mathtt{Goal Stack}, \mathtt{Proven Goals}, \mathtt{Staging Output} >
	\)
	\caption{Prover State}
	\label{fig:ProverState}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}
\subsection{Input and output handling}
To understand how our prover handles input and output, it is important to note that the flow of the LOOP-MODE module is as follows: 
\begin{enumerate}
	\item User input is placed in the \(\mathtt{INPUT}\) member of the loop state
	\item Maude engine applies all the rewritings possible, resulting in a new loop state.

@@

\subsection{Prover execution flow}
In this section we will describe the flow of a demonstration made using our prover. The most important parts of the prover's execution are controlled through the first member of a prover state, \texttt{Action}, which specifies the next step in the processing of the current goal. There are two types of actions, \texttt{System Actions}, colored in the figure in purple and \texttt{User Actions}, colored in light-yellow. The diagram presenting how the prover works is presented in Figure \ref{fig:ProverDiagram}
\\
\begin{enumerate}
	\item The user issues the command \texttt{loop init .} in the Maude console. This will initialize the loop state.
	\item Now the user will have to specify the goal it wants to prove, by issuing the command \begin{center}(\texttt{prove(G)})\end{center}, where G is the representation of the goal using the syntax defined for our logics. The command will populate the required fields of the prover state (\texttt{RootGoal}, \texttt{GoalStack}) with G.
	\\

	This command will also populate the \texttt{ProvenGoals} list with axioms.
	\\

@@

	\item The next phase involves applying the logic proof rules to the goals in the \texttt{GoalStack} to generate sub-goals and prove them. Once a goal is popped from the \texttt{GoalStack}, it can be replaced by sub-goals generated by proof rules, matched with an proven goal / axiom, be proven automatically or be admitted manually. This phase lasts until there are no more goals left in the \texttt{GoalStack}, meaning that the \texttt{RootGoal} has been proven.

@@

\paragraph{Goal Graph} One of the improvements which could make the user experience much better for our prover would be the capability to backtrack in the proof and try another approach for the same goal. The initial idea for this prover was to have a \texttt{GoalGraph} instead of \texttt{GoalStack} such that when the user applies a rule to a goal, the sub goals are represented as nodes linked to the goal from which they originated, making it possible to trace back the progress to intermediary states. This approach would also make the \texttt{ProvenGoals} a lot more useful by increasing the chance of running into an already proven goal and the ability to mark complex goals as proven. At the moment, the only goals which can be used from the \texttt{ProvenGoals} are those admitted manually by the user, since the other proven goals already matched axioms.

@@

		[\texttt{SL}] + \texttt{apply} + \textit{RuleName}

@@

		With \texttt{SL} being added only when applying Separation Logic rules, to distinguish between the Relational Separation Logic rules with the same name.

@@

			\item If the result corresponds to what the prover was expecting, we modify the state by adding the newly generated goals to the \texttt{GoalStack}, by moving down the "meta"-hierarchy - parsing meta-representation of terms.
			\item The last thing to note is an example of how the rules generate output independent of other applications. In the last field of the \texttt{Prover State}, as \texttt{StagingOutput}, there will be a list of quoted identifiers which will eventually get moved to the loop state and then outputted to the user.

@@

	\begin{lstlisting}[label=axiom,caption=Initialization of Prover state]
rl [proveGoal] : < prove(HQ) ; RG ; G ; GS ; GL ; nil > =>
< noAction ; HQ ; noGoal ; HQ <> emptyGL ; 
downTerm('`{_`}_`{_`}['_|->_['E:AExp,'_:AExp],'`[_`]:=_['E:AExp,'F:AExp],'_|->_['E:AExp,'F:AExp]] , emptyGL)
--- ... [other axioms omitted for clarity] 
; 'begin '\n > .\end{lstlisting}
	\item Before attempting to match terms we also interpret the equality between variables using the \texttt{replace} operator introduced previously to reduce the \texttt{Goal} to a simpler form, which might match to an axiom.
	\begin{lstlisting}[caption=Equation reducing the goal before attempting to match it]]
ceq anyMatches(G, G1 <> GL) = G1
---if G := G1 .
if metaXmatch(upModule('PROVER-INTERFACE, false), 
upTerm(G1),
--- Try to reduce G as much as possible before matching
getTerm(metaRewrite(upModule('SYMBOLIC-EXECUTION, false), upTerm(G), unbounded)),
nil, 0, unbounded, 0) :: MatchPair .
	\end{lstlisting} 
	\begin{lstlisting}[caption=Rule intepreting variable equality]
rl {P * (emp /\ X eqs Y)} X1 := E {Q} =>  
if X1 =/= X then 
{replace(P, X, Y)} replace(X1 := E, X, Y) {replace(Q, X, Y)}
else 
{replace(P, X, Y)} X1 := replace(E, X, Y) {Q}
fi .\end{lstlisting}
\end{itemize}
\subsubsection{Automatic demonstration of implications}
Automatic demonstration of implications is one of the most challenging aspects of this paper and this section will present our incipient approach to this problem.

We have modeled implication proof rules as rewrite rules in Maude (example in Listing \ref{implicationProofRule}), in order to make use of its \texttt{metaSearchPath} operator, which will search for a series of rewrites which transform a term into another.
\\
We used this functionality to try and solve implications 
\[
	R \Rightarrow S
\]
by searching for a series of rewrites from \texttt{R => S} to \texttt{true}. 
\\ 
\begin{lstlisting}[label=implicationProofRule,caption=Example of implication rewrite rules]
rl R:Relation * ( S:Relation \/ T:Relation ) => (R:Relation * S:Relation) \/ (R:Relation * T:Relation) .
--- Example for the usage of the replace operator
rl R * (Emp /\ X eqs Y) => replace(R, X , Y) .
rl (R:Relation => R:Relation) => true .
\end{lstlisting}
Most proof rules are of the form \(R \Leftrightarrow S\) and thus can be naturally expressed in our prover using rewrite rules. However, when the proof rules involve just a one way implication, such as \(Emp \Rightarrow Same\), if we were to model it simply as the first rule in the Listing \ref{implicationProofRule}, an implication \texttt{Same => Emp} would be considered true, because Maude would rewrite the \texttt{Emp} to \texttt{Same} and \texttt{Same => Same} evaluates to \texttt{true}. To prevent this from happening, we have constrained one-way implications by providing the entire context in which they can be rewritten :
\begin{lstlisting}
rl ((R * Emp) => S) => ((R * Same) => S) .
\end{lstlisting}

We have opted for the approach to search from the term \texttt{R => S} to \texttt{true} instead of searching from \texttt{R} to \texttt{S} because there are implications in which it is beneficial to use information from both sides in order to prove it. A simple example would be \texttt{R => true} which should evaluate to \texttt{true} - the rewrite rule corresponding to this implication is used in the proof at Appendix \ref{proofExample} - but if we were to search from \texttt{R} to \texttt{true}, the search will probably fail.

@@

The second parameter controls the depth up to which the search goes. The prover will try automatically to search a solution within a depth of 3 rewrites, but the user can manually specify searching for a solution using the \texttt{auto x} command, where \texttt{x} is the depth up to which Maude should try to search the solution (our experiments show that searches up to a depth around 12 should finish use a reasonable amount of resources and finish in under 10 minutes - this observation of course depends on the implication and the number of possible rewrites starting from it). The found path is presented to the user if the implication has been proven. 
\\

In the future, one might consider using Maude internal strategies \cite{manual} to control the rewriting and searching in the implication proving process, to make it follow heuristics and thus be more effective.

\paragraph{Inductive predicates} - the prover currently supports defining inductive predicates only by modifying the underlying Maude code. We have defined a single predicate, used in the Relational Separation Logic paper \cite{relational} :
\begin{lstlisting}[caption=Predicate syntax definition]
op List : Id -> Relation . 
\end{lstlisting}
In terms of the semantics of the \texttt{List} predicate, we have added rules to the \textbf{implication-prover.maude} file, to be considered when rewriting implications.
\begin{lstlisting}[caption=Example of rule interpreting the List predicate during implication proving]
rl [ (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ; (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ] * List(Y:Id) => List(X) .
\end{lstlisting}
\paragraph{Cyclist}\cite{cyclist} \cite{cyclistSite} is an automatic entailment prover for Separation Logic which we planned to integrate with our prover for Separation Logic implications and eventually extend to Relational Separation Logic, since it provides a generic framework. Since Maude does not easily support external commands, the idea would be to modify Maude's C++ code and add a hook for the \texttt{[special]} operator attribute which would invoke that hook, which will call the external command for Cyclist. We have however focused on other aspects of our prover, and left this feature for the future.
\subsection{User Interface}
Since our project aims to simplify the demonstrations of Separation Logic / Relational Separation Logic formulae, using an interactive prover, the user interface plays a major role in the success of the project - the easier it is to use and understand, the more chances of success it has. This idea has been presented in \cite{primer}, when discussing key elements of tools related to program verification. \\

@@

Maude has been interesting to work because it required an entirely different mindset than the mainstream programming languages. Once grasping the concepts around Maude which was built, it was easy to come up with simple, clever solutions to the encountered problems. \\
As it is described on it's website \cite{maudesite}, Maude is a powerful logical framework which allows modeling of other logics in it; it proved to be exactly that, enabling us to represent the logics in a natural manner, which made the implementation much less error-prone and allowed us to focus on the relevant part of this project, namely how to apply the logics instead of focusing on how to properly implement them.
\\

@@

\begin{appendices}
	\section{Example of proof using the prover}
	\label{proofExample}
	This appendix contains all the necessary commands to be run in a Maude console in order to run an example demonstration using the prover.
\begin{lstlisting}
load prover-interface.maude
loop init .
(prove(
{ 
((
Same 
*
List(c) 
*
[ y |-> x0 ; yp |-> x0 ]
) 
*
(y eqs yp /\ Emp))
*
(c eqs cp /\ Emp)
}
pass ; 
while (! (c eqs null))
do 
x := [y] ; 
[c] := 0 minus x ;
c := [c plus 1]
od
--
xp := [yp] ;
while (! (cp eqs null))
do 
[cp] := 0 minus xp ;
cp := [cp plus 1] 
od
{
(Same
*
(y eqs yp /\ Emp))
* 
(c eqs cp /\ Emp)
}
)
)
(applySequencing ( Same * List(c) * [ y |-> x0 ; yp |-> x0 ] * (y eqs yp /\ Emp) * (c eqs cp /\ Emp) * (xp eqs x0 /\ Emp) ) )
(applyFrameRule (Same * List(c) * (y eqs yp /\ Emp) * (c eqs cp /\ Emp)) )
(
applyConsequence
( [ (y |-> x0) ; (yp |-> x0) ] )
--
( [ (y |-> x0) ; ((yp |-> x0) * (xp eqs x0 /\ emp)) ] )
)
(applyEmbeddingRule)
(
applyConsequence
( Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0] )
--
( (Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0]) /\ !(!(c eqs null)) ) --- R si not B
)
(applyLoop)
(applyConsequence
([(c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c0 /\ emp) 
; (c0 |-> -) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c0 /\ emp) ] 
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
--
([(c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c1 /\ emp) 
; (c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c1 /\ emp) ]
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
)
(ok) 
(applyFrameRule (List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0]))
(applyEmbeddingRule)
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0)))
(SLapplySequencing ((c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0) * (emp /\ x eqs x0)))
(SLapplyFrameRule ((c0 |-> -) * ((c0 plus 1) |-> c1) * (emp /\ c eqs c0)))
(SLapplyFrameRule ( (y |-> x0) * ((c0 plus 1)|-> c1) ) )
(SLapplyFrameRule ( (y |-> x0) *  (c0 |->(0 minus x0)) ) )
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp) * (emp /\ cp eqs c0)))
(SLapplyFrameRule ( (c0 plus 1)|-> c1 ))
(SLapplyFrameRule  ((emp /\ (xp eqs x0) ) * (c0 |->(0 minus x0))) )
(ok)
(auto 9)
\end{lstlisting}
\end{appendices}
